Competitive differential chemotaxis can explain the self-organisation of retinotopic maps

### 2023-08-25 paper/main.tex
Updates paper from changes made on Overleaf in Jan 2023

\title {
Competitive differential chemotaxis can explain the self-organisation of topographic maps
}

@@

Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}.
%
The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'. 
The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones. 
%
This key idea (called chemospecificity) does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target.  
An additional component of Sperry's theory \citep{stevens_handbook_1950,sperry_problems_1955,sperry_chemoaffinity_1963} suggested that the axon could determine this direction by evaluating local information in gradients of chemicals spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.

Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity \citep{oleary_molecular_1999}.
In this system a two-dimensional map of  the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum. 
This retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum).
This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

In the 1990s, the theory of chemoaffinity and chemotaxis was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}. % And sounds? 
%
The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo studies \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002}.
Two classes of ephrins have been identified, labelled `A' and `B'.
Ephrin A and its receptors mediate a repulsive interaction which contributes to an ordering of RGC axons on the tectum along the rostral-caudal axis. Ephrin B acts on medial-lateral ordering with an attractive interaction. 
Together, the ephrins are able to provide a coordinate system on the tectum to guide growing axons, but precisely how this positional information is used by axonal growth cones is not yet fully understood.

Quantitative models exploring axonal ordering were already in existence by the time of the discovery of the ephrins. 
The importance of chemical gradients to biological processes was already well known from the study of slime moulds \citep{bonner_evidence_1947}. 
Sperry himself discussed chemical gradients in relation to the problem of neuronal pathfinding \citep{sperry_problems_1955}, albeit qualitatively.  
Some time after Sperry's active period, but before the ephrins were identified, \citet{gierer_development_1981,gierer_model_1983,gierer_directional_1987} made one of the few quantitative investigations of a gradient-based model of the retinotectal projection. He presented gradient following as a special case of a general class of \emph{potential models}. 
In a potential model, the interactions between the properties of a growing axon originating from a location $\mathbf{u}$ on the retina and its environment at location $\mathbf{x}$ on the tectum set up some potential function, $p(\mathbf{x},\mathbf{u})$. 
The axon then grows or moves to minimize $p$, which causes it to find its target location. This model successfully creates the wildtype map because each axon originates with a different $\mathbf{u}$ and has its minimum $p(\mathbf{x},\mathbf{u})$ in the correct target location.
This model forms a mathematical expression of Sperry's theory. Because it is quantitative, it allowed Gierer to identify two features of the theory: i) A \emph{pair} of opposing gradients is needed to form a potential function in one dimension (and \emph{two} pairs to create a 2D potential function). ii) Tectal or retinal ablation would require $p$ to be re-specified to reproduce the experimental results that fibres from an ablated retina will generate a map across the entire tectum and fibres from an intact retina will form a compressed map on an ablated tectum \citep{attardi_preferential_1963,schmidt_retinal_1978,schmidt_expansion_1978}.
Gierer addressed the second problem by introducing a form of competition between fibres, postulating that $p$ would accumulate locally at a rate proportional to the fibre density in that location and he noted that this was equivalent to a more direct fibre-fibre competition.
He demonstrated, in a one-dimensional computer simulation, that this form of competition would lead to the expected response to ablations \citep{gierer_model_1983}.

Gierer's simulations confirmed the validity of this basic idea, which we will refer to as \emph{differential chemotaxis}, for the problem of map formation with respect to a single spatial dimension \citep{gierer_model_1983,karbowski_model_2004}. 
The model helped establish how \emph{competition between axons} for space in the target tissue can additionally account for  map development after tissue ablation.

Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}. 
These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue. 
In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples). 
However, we found it numerically intractable to extend the model described in \citet{james_modelling_2020}, which simulated 41 whisker-specific projections, to the retinotectal system for which hundreds of projections are required (to demonstrate surgical manipulations).
Instead, we sought to implement an \emph{agent-based} model of differential chemotaxis, in which each axon growth cone is an agent that interacts with its environment and with the other growth cone agents in the system.
With this model, we asked: i) Can a two-dimensional implementation of Gierer's model explain the full range of surgical manipulations that have been described in the literature? ii) Can the model be developed such that only \emph{local} interactions are required to transfer information between agents? and iii) Can a differential chemotaxis model explain more recent results in which levels of EphA receptor sub-types are changed by genetic manipulation \citep{brown_topographic_2000,reber_relative_2004}?

@@

All data and code required to reproduce the results in this work are available in the public repository 
\url{https://github.com/sebjameswml/RetinoTectal}.
The software was built using the library morphologica \citep{james_morphologica_2022}.

>>>>

\title {
Competitive differential chemotaxis can explain the self-organisation of retinotopic maps
}

@@

Since the pioneering work of Sperry \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}, progress on understanding topographic patterning of brain connectivity has been made through a synergy of experimental and computational neuroscience, focussed on the projection of retinal ganglion cell axons into the optic tectum. Since computers first became available as scientific tools, a wide range of theories have been developed and tested through formal algorithms and computer simulation work. Models can be distinguished on a number of grounds, such as the level of abstraction at which they describe the process of map development, and the assumptions they represent about the contributions of chemotaxis, molecular interactions between axons, and competition for space in the destination tissue (for reviews see \citealp{Prestige1975,Goodhill2005}).

One important class of theories assumes that map formation is driven by chemotaxis, influenced by pairs of anti-parallel expression gradients in the tectum, with the balance of influences in each axon determined by a receptor density specific to its origin in the retina (\citealp{Gierer1983,Gierer1987,Simpson2011}; see also \citealp{Karbowski2004,James2020}). A second class of theories assumes that a single gradient is primarily responsible for chemotaxis and that maps reflect additional processes such as competition \citep{Triplett2011}, Hebbian activity \citep{Tsigankov2006,Tsigankov2010}, or marker induction \citep{Prestige1975,Willshaw2006}.

A key test of the validity of recent models is their ability to account for data from a set of genetic manipulation studies by \cite{brown_topographic_2000} and \cite{reber_relative_2004}, in which axons from a subset of retinal ganglia cells treated with a Eph-A3 knock-in form a separate, rostrally shifted topographic map, which becomes indistinct from that formed by untreated cells, at a ‘collapse point’ on the rostral side. \cite{Sterratt2013} found that a mixture of counter-gradients and an additional competition term provided the best account of these data. A follow-up study in which a range of different computational models were systematically compared \citep{hjorth_quantitative_2015} found that one focused on competition by \cite{Triplett2011} could best account for several such experimental findings, and that the other types of model could not explain the collapse point.

The model of \cite{Triplett2011} describes maps as minima of an energy function, whose structure can be obtained via a stochastic optimisation process. Accordingly, the energy is the sum of three terms, relating to the influences of chemical gradients, axon density (competition), and activity-dependent growth processes. Maps that optimise this function can be obtained by starting with a random map and then substituting connections at random, accepting those substitutions with a probability that increases if they reduce the energy, and repeating until the algorithm converges on a pattern of connections corresponding to a minimal energy state.

The ability of this model to explain the collapse point in the Eph-A3 knockin studies can be understood in terms of its relationship to simpler models developed previously, with which it shares (via the activity-dependent term in the energy function) similar assumptions \citep{Koulakov2004}. According to this earlier model, pairs of adjacent axons on the tectum are evaluated iteratively and the probability of their positions being switched increases with an energy defined by the product of the difference in Eph receptor density and the difference in ephrin ligand density. As the retinal receptor and tectum ligand densities express anti-parallel gradients, a positive energy is translated into a pairwise switch of axon positions with high probability. Eph knock-ins are assigned higher receptor density values, and the map collapse point emerges when differences in receptor density (due to retina position and treatment condition) are offset by differences in ephrin ligand density in the tectum. The two terms are balanced at a point that shifts further in the direction of cells with a more temporal retina origin for a stronger (i.e., homozygous) knock-in, as observed experimentally.

These models inherit from thermodynamic models used to investigate phase transitions in statistical physics, the concept of a `temperature', via a parameter that weights the switching probabilities and effectively determines the sharpness of the collapse between treated and untreated maps. Interestingly, at high temperatures (sharp switching) the model of \cite{Koulakov2004} becomes formally equivalent to the `arrow’ model, one of the first models of retinotectal map self-organisation proposed  \citep{Hope1976}.

The explanatory power of these `switching' models stems from their simplicity and degree of abstraction, revealing the map pattern that optimises the constraints represented by a given energy function. However, while switching models are evaluated through a numerical procedure (i.e., iterative switching), that procedure should not necessarily be interpreted as an explicit representation of the process of development \citep{Wilson2015}. Consider, for example, that from more realistic initial conditions than the uniform random map typically assumed, local energy minima quite unlike organised biological maps are more likely to be encountered.

In a parallel strand of work the principles represented by switching models have been explored in models that represent the developmental process more explicitly. For example, \cite{Overton1982} showed how the net effects of iterative switching in discrete systems could be achieved via a continuous dynamics based on iteratively ‘nudging’ axons by a velocity vector that is weighted by the factors considered by the arrow model, including for example, functions of the proximity of growing axons to their neighbours.

Continuing this strand of work, \cite{Simpson2011} have more recently extended the model of \cite{Overton1982} from its original description of axon growth in one spatial dimension to simulate growth in two dimensions, and they have shown that it can account for a wide range of data from genetic studies, as well as a wide range of surgical manipulation experiments; including the emergence of a map collapse point in simulated Eph-A3 knock-in experiments.

The models of \cite{Overton1982} and \cite{Simpson2011} incorporate three basic ingredients; competition for space, receptor-based interactions between axons, and gradient-based guidance. However, a major limitation is that chemotaxis is not implemented explicitly. Instead, simulations involve providing growing axons with explicit directional information about their target destination at every timestep. This was an appropriate choice given the scientific goals of \cite{Simpson2011}, but for this reason the theory of retintopic map formation that is represented by their model cannot be tested comprehensively, and indeed this model was excluded from the systematic comparison of models by \cite{Hjorth2015} on that basis.

An open question then, which we address in this paper, is whether the \cite{Simpson2011} model can still account for a wide range of experimental data from surgical and genetic experiments, when its assumptions about chemotaxis are represented explicitly, i.e., when simulated axon growth is guided by only local information? If so, such a model would represent the most complete account of retinotectal map self-organization to date.

%Sperry's chemoaffinity model is an influential theory of how axons from neurons that originate in one tissue can find their way to specific locations in a target tissue during early brain development \citep{sperry_reestablishment_1942,sperry_visuomotor_1943,sperry_chemoaffinity_1963}.

%The central proposal of the theory is that growing axons find their location in a destination tissue by means of `labels of a cytochemical nature'. The chemical affinity between labels in source and destination tissues allows axons to distinguish correct target locations from incorrect ones.

%This key idea (called chemospecificity) does not, on its own, explain how an axon can find the correct \emph{direction} in which it should move to reach its target. An additional component of Sperry's theory \citep{stevens_handbook_1950,sperry_problems_1955,sperry_chemoaffinity_1963} suggested that the axon could determine this direction by evaluating local information in gradients of chemicals spanning the target tissue, giving rise to chemotaxis of the axonal growth cones.

%Much of Sperry's work was carried out on the retinotectal system, which is ideal for the study of ordered axonal development in fish, amphibia and birds, because in these animals, patterning of axons across the tectum is largely complete prior to the onset of neural activity \citep{oleary_molecular_1999}.

%In this system a two-dimensional map of  the retina is developed on the surface of the optic tectum such that the axons from adjacent retinal ganglion cells (RGCs) find their way to adjacent locations on the tectum.  This retinotectal map is amenable to a number of surgical manipulations especially in amphibia, in which re-growth of the projection can be observed after surgical changes to the sensory apparatus (such as ablation or rotation of the retina) or the map substrate (the tectum). This allowed Sperry and other researchers to carry out a wide range of experimental manipulations, the results of which informed the chemoaffinity theory.

%In the 1990s, the theory of chemoaffinity and chemotaxis was given robust support by the discovery of the ephrin ligands and their receptors \citep{cheng_complementary_1995,drescher_vitro_1995} which have been shown to form into graded expression fields in the retina \citep{braisted_graded_1997}, tectum \citep{braisted_graded_1997,feldheim_genetic_2000} as well as in other sensory systems, such as the somatosensory system \citep{vanderhaeghen_mapping_2000}. % And sounds?

%The Ephrin ligands have a clear effect on axonal outgrowth, as shown in-vitro \citep{cheng_complementary_1995,drescher_vitro_1995,hansen_retinal_2004} and by in-vivo studies \citep{frisen_ephrin-a5_1998,rodger_transient_2000,mann_topographic_2002,hindges_ephb_2002}. Two classes of ephrins have been identified, labelled `A' and `B'. Ephrin A and its receptors mediate a repulsive interaction which contributes to an ordering of RGC axons on the tectum along the rostral-caudal axis. Ephrin B acts on medial-lateral ordering with an attractive interaction. Together, the ephrins are able to provide a coordinate system on the tectum to guide growing axons, but precisely how this positional information is used by axonal growth cones is not yet fully understood.

%Quantitative models exploring axonal ordering were already in existence by the time of the discovery of the ephrins.  The importance of chemical gradients to biological processes was already well known from the study of slime moulds \citep{bonner_evidence_1947}.  Sperry himself discussed chemical gradients in relation to the problem of neuronal pathfinding \citep{sperry_problems_1955}, albeit qualitatively. Some time after Sperry's active period, but before the ephrins were identified, \citet{gierer_development_1981,gierer_model_1983,gierer_directional_1987} made one of the few quantitative investigations of a gradient-based model of the retinotectal projection. He presented gradient following as a special case of a general class of \emph{potential models}. In a potential model, the interactions between the properties of a growing axon originating from a location $\mathbf{u}$ on the retina and its environment at location $\mathbf{x}$ on the tectum set up some potential function, $p(\mathbf{x},\mathbf{u})$.  The axon then grows or moves to minimize $p$, which causes it to find its target location. This model successfully creates the wildtype map because each axon originates with a different $\mathbf{u}$ and has its minimum $p(\mathbf{x},\mathbf{u})$ in the correct target location. This model forms a mathematical expression of Sperry's theory. Because it is quantitative, it allowed Gierer to identify two features of the theory: i) A \emph{pair} of opposing gradients is needed to form a potential function in one dimension (and \emph{two} pairs to create a 2D potential function). ii) Tectal or retinal ablation would require $p$ to be re-specified to reproduce the experimental results that fibres from an ablated retina will generate a map across the entire tectum and fibres from an intact retina will form a compressed map on an ablated tectum \citep{attardi_preferential_1963,schmidt_retinal_1978,schmidt_expansion_1978}.

%Gierer addressed the second problem by introducing a form of competition between fibres, postulating that $p$ would accumulate locally at a rate proportional to the fibre density in that location and he noted that this was equivalent to a more direct fibre-fibre competition. He demonstrated, in a one-dimensional computer simulation, that this form of competition would lead to the expected response to ablations \citep{gierer_model_1983}.

%Gierer's simulations confirmed the validity of this basic idea, which we will refer to as \emph{differential chemotaxis}, for the problem of map formation with respect to a single spatial dimension \citep{gierer_model_1983,karbowski_model_2004}.  The model helped establish how \emph{competition between axons} for space in the target tissue can additionally account for  map development after tissue ablation.

%Simulations of differential chemotaxis and axonal competition have recently been extended to investigate map development in two spatial dimensions, revealing how these two ingredients can give rise to realistic patterns of synaptic innervation \citep{james_modelling_2020}.  These models are formulated as systems of partial differential equations (PDE), and solved through numerical integration on an underlying lattice structure representing the target tissue.  In general, this approach makes mathematical analysis of system dynamics tractable, and as such PDE modelling has played a central r\^ole in the development of theories of pattern formation (see \citet{krause_modern_2021} for recent examples).  However, we found it numerically intractable to extend the model described in \citet{james_modelling_2020}, which simulated 41 whisker-specific projections, to the retinotectal system for which hundreds of projections are required (to demonstrate surgical manipulations). Instead, we sought to implement an \emph{agent-based} model of differential chemotaxis, in which each axon growth cone is an agent that interacts with its environment and with the other growth cone agents in the system. With this model, we asked: i) Can a two-dimensional implementation of Gierer's model explain the full range of surgical manipulations that have been described in the literature? ii) Can the model be developed such that only \emph{local} interactions are required to transfer information between agents? and iii) Can a differential chemotaxis model explain more recent results in which levels of EphA receptor sub-types are changed by genetic manipulation \citep{brown_topographic_2000,reber_relative_2004}?

@@

All data and code required to reproduce the results in this work are available in the public repository 
{https://github.com/sebjameswml/RetinoTectal}.
The software was built using the library morphologica \citep{james_morphologica_2022}.
